Animacy Detection in Stories

### 2014-12-17 paper/paper.tex
update paper

A common view supported by philosophers such as Daniel Dennett is that
animacy and aliveness are better to be treated as epistemological
stances rather than fixed states in the world \cite{dennett:96}: not
ineffable qualia but behavioral capacity defines our stance towards
objects. Depending on whether people \textit{think} that an object is
animate, they utilize different cognitive strategies to explain and
predict the actions of those objects. Th following opening of a
well-known story will make this point more clear:

@@

Given the fundamental nature of the distinction between animate and
inanimate, it is perhaps not too surprising that it has proven to be
useful in a variety of natural language processing tasks dealing with
e.g.\ anaphora resolution and dependency parsing
\cite{orasan:07,lee:13,ovr:niv:07}. Existing methods to automatically
label natural language for animacy are usually either rule based or
apply machine learning techniques, or make a combination of
both. Common to most approaches is that they make use of semantic
lexicons with information about animacy and syntactic cues in a
text. Both feature types are relatively costly to obtain as they
require large vocabularies or syntactic parsing systems, which, with
the exception of a few languages, are not readily available.

In this paper we present a new linguistically uninformed model to
automatically label texts for animacy. We will show that we can (mostly)
do away with features that require syntactic parsing or semantic
lexicons while still yielding competitive performance. We will focus
on labeling animacy in stories because stories pose some interesting
problems to automatic systems of animacy recognition. As the example
of the fleeing pancake already made clear, quite some entities in
stories, though inanimate in the `real' world, exhibit all behavioral
capacities to consider them to be animate and rational agents. Such
examples make a clear case for developing dynamic, data driven systems
that do not rely too much on static and fixed world knowledge.

@@

\cite{ovrelid:04,ovrelid:05,ovrelid:06,ovrelid:08,ovrelid:09} present
a number of animacy classifiers that make use of syntactic and
morphological features. These features include the frequency of
analysis of the noun as `subject' or `object', the frequency of the
occurence of a noun in a passive $by$-phrase, and the frequency of the
noun as a subject followed by either animate personal pronouns or
inanimate personal pronouns. These features are then aggregated for
each lemma after which a machine learning system (decision tree or
$k$-nearest neighbor classifier) is trained. In \cite{bowman:12} a
similar approach is presented. In this study a Maximum Entropy
classifier is trained on the basis of three feature types: (1)
bag-of-words with and without their corresponding Part-of-Speech tags,
(2) internal syntactic features such as the syntactic head and (3)
external syntactic features that describe the dependency relation of a
noun to a verb (i.e.\ subject relation, object relation etc.)  This is
the only study that makes use of a fully labeled corpus for animacy.
Not so much concerned with animacy detection, \cite{karsdorp:12}
attempt to extract the cast (i.e.\ all characters) from a
story. Similar to \cite{bowman:12} they rely on dependency tags to
extract the subjects of direct and indirect speech.

@@

% commentary on above approaches
The approaches mentioned above present us with a number of
problems. First, almost all of them heavily rely on costly,
linguistically informed features derived from lexical-semantic
databases or syntactic parsing. For most languages in the world,
however, we cannot rely on these resources, either because they do not
exist, or because their performance is insufficient.  Second, animacy
detection is often seen as a useful feature for a range of natural
language processing techniques, such as anaphora resolution and
syntactic parsing. The mutual dependence of these techniques and
animacy detection, however, leads to a chicken and egg
situation. Another problem with the approaches above is that they are
lemma-based, which means that the models are generally insensitive to
different usages of a particular word. A pancake is typically not an
animate entity, but, as we have seen, whether it is animate or not may
depend on the story in which it appears, and even within the story the
animacy may change. We therefore need to develop a dynamic, data
driven system that is token based and does not rely too much on static
and fixed world knowledge.

@@

\subsubsection{Semantic Features} The most innovative feature we have
included in our model is concerned with semantic similarity. In his
\textit{Philophische Untersuchungen} Wittgenstein already suggests
that ``Die Bedeutung eines Wortes ist sein Gebrauch in der Sprache''
(PI 43). The well-known insight in computational linguistics that the
meaning of words can be approximated by comparing the linguistic
contexts in which words appear, is an implementation of this very
idea. The idea is very simple, and equally strong: words that often
co-appear with the same set of words, will have a more similar
meaning. Recently, there has been a lot of interest in procedures that
can automatically induce so-called `word-embeddings' from large,
unannotated collections of texts. These models typically attempt to
learn a lower-dimensional vector for each word in the vocabulary which
captures the typical co-occurrence patterns of the word in the
corpus. The similarity between words can then be approximated by
applying arithmetics to these vectors of word embeddings.

@@

We add the word embedding vectors to our animacy model as follows. We
augment each word in the data with its corresponding 300 dimension
word embeddings vector. Out-of-vocabulary words were given a vector were all dimensions are set to zero.

@@

In table \ref{tab:results-all} we present the results for all nine
models on the complete data set. For each model we report the
precision, recall, $F1$-score and Area Under the Precision-Recall
Curve (AUC) for the animate words and the
inanimate words.

\begin{table}
\centering
\begin{tabular}{llrrrr}
\toprule
features & class & precision &  recall & $F1$ &  AUC \\
\midrule
\multirow{2}{*}{embeddings}                            & inanimate     &       0.98 &    0.99 &    0.98 & \multirow{2}{*}{0.93} \\
                                                       & animate     &       0.92 &    0.89 &    0.91 &  \\
\multirow{2}{*}{word}                                  & inanimate     &       0.96 &    0.99 &    0.98 & \multirow{2}{*}{0.94} \\
                                                       & animate     &       0.94 &    0.78 &    0.85 &  \\
\multirow{2}{*}{word + embeddings}                     & inanimate     &       0.98 &    0.99 &    0.98 & \multirow{2}{*}{0.96} \\
                                                       & animate     &       0.93 &    0.90 &    0.91 &  \\
\multirow{2}{*}{word + PoS}                            & inanimate     &       0.97 &    0.99 &    0.98 & \multirow{2}{*}{0.96} \\
                                                       & animate     &       0.94 &    0.86 &    0.90 &  \\
\rowcolor{Gray}                                        & inanimate     &       0.98 &    0.99 &    0.99 &  \\
\rowcolor{Gray}\multirow{-2}{*}{word + PoS + embeddings}   & animate     &       0.94 &    0.92 &    0.93 & \multirow{-2}{*}{0.97} \\
\multirow{2}{*}{word + PoS + lemma}                    & inanimate    &       0.98 &    0.99 &    0.98 & \multirow{2}{*}{0.96} \\
                                                       & animate     &       0.94 &    0.87 &    0.90 &  \\
\multirow{2}{*}{word + PoS + lemma + embeddings}       & inanimate     &       0.98 &    0.99 &    0.99 & \multirow{2}{*}{0.97} \\
                                                       & animate     &       0.94 &    0.92 &    0.93 &  \\
\multirow{2}{*}{word + PoS + lemma + dep}              & inanimate     &       0.98 &    0.99 &    0.98 & \multirow{2}{*}{0.96} \\
                                                       & animate     &       0.94 &    0.88 &    0.91 &  \\
\multirow{2}{*}{word + PoS + lemma + dep + embeddings} & inanimate     &       0.98 &    0.99 &    0.99 & \multirow{2}{*}{0.97} \\
                                                       & animate     &       0.94 &    0.92 &    0.93 &  \\
\bottomrule
\end{tabular}
\caption{Precision, recall, $F1$ and Area Under the Curve (AUC) scores for animate and inanimate classes per feature setting for all words.}
\label{tab:results-all}
\end{table}

All models perform well on classifying inanimate words. However, since
this is the majority class, it is of course far more interesting to
compare the performance of the models on the animate instances. It is
interesting to observe that the `simple' $n$-gram word model already
performs rather well. Adding more features, like Part-of-Speech,
lemmata, etcetera, only has a positive impact on the recall of the
model, while leaving the precision untouched. As can be observed from
the table, employing the rather expensive dependency features shows
barely any improvement.

@@

\begin{table}
\centering
\begin{tabular}{llrrrr}
\toprule
features & class &  precision &  recall & $F1$ &  AUC \\
\midrule
\multirow{2}{*}{embeddings}                            & inanimate     &       0.89 &    0.94 &    0.92 & \multirow{2}{*}{0.92} \\
                                                       & animate       &       0.91 &    0.85 &    0.88 &  \\
\multirow{2}{*}{word}                                  & inanimate     &       0.78 &    0.98 &    0.87 & \multirow{2}{*}{0.91} \\
                                                       & animate       &       0.96 &    0.60 &    0.74 &  \\
\multirow{2}{*}{word + embeddings}                     & inanimate     &       0.90 &    0.96 &    0.93 & \multirow{2}{*}{0.96} \\
                                                       & animate       &       0.94 &    0.86 &    0.89 &  \\
\multirow{2}{*}{word + pos}                            & inanimate     &       0.86 &    0.96 &    0.90 & \multirow{2}{*}{0.94} \\
                                                       & animate       &       0.92 &    0.78 &    0.84 &  \\
\rowcolor{Gray}                                        & inanimate     &       0.93 &    0.96 &    0.94 &  \\
\rowcolor{Gray}\multirow{-2}{*}{word + PoS + embeddings} & animate       &       0.94 &    0.89 &    0.92 &\multirow{-2}{*}{0.97}  \\
\multirow{2}{*}{word + PoS + lemma}                    & inanimate     &       0.87 &    0.96 &    0.91 & \multirow{2}{*}{0.95} \\
                                                       & animate       &       0.94 &    0.80 &    0.86 &  \\
\multirow{2}{*}{word + PoS + lemma + embeddings}       & inanimate     &       0.93 &    0.96 &    0.94 & \multirow{2}{*}{0.97} \\
                                                       & animate       &       0.94 &    0.89 &    0.92 &  \\
\multirow{2}{*}{word + PoS + lemma + dep}              & inanimate     &       0.87 &    0.96 &    0.91 & \multirow{2}{*}{0.95} \\
                                                       & animate       &       0.93 &    0.80 &    0.86 &  \\
\multirow{2}{*}{word + PoS + lemma + dep + embeddings} & inanimate     &       0.93 &    0.96 &    0.94 & \multirow{2}{*}{0.97} \\
                                                       & animate       &       0.94 &    0.90 &    0.92 &  \\
\bottomrule
\end{tabular}
\caption{Precision, recall, $F1$ and Area Under the Curve (AUC) scores for animate and inanimate classes
  per feature settings for all words tagged as noun.}
\label{tab:results-noun}
\end{table}

>>>>

A common view supported by philosophers such as Daniel Dennett is that
animacy and aliveness are to be treated as epistemological
stances rather than fixed states in the world \cite{dennett:96}: not
ineffable qualia but behavioral capacity defines our stance towards
objects. Depending on whether people \textit{think} that an object is
animate, they utilize different cognitive strategies to explain and
predict the actions of those objects. Th following opening of a
well-known story will make this point more clear:

@@

Given the fundamental nature of the distinction between animate and
inanimate, it is perhaps not too surprising that it has proven to be
useful in a variety of natural language processing tasks dealing with
e.g.\ anaphora resolution and dependency parsing
\cite{orasan:07,lee:13,ovr:niv:07}. Existing methods for the automatic
labeling of text for animacy are usually either rule based or apply
machine learning techniques, or make a combination of both. Common to
most approaches is that they make use of semantic lexicons with
information about animacy and syntactic cues in a text. Both feature
types are relatively costly to obtain as they require large
vocabularies or syntactic parsing systems, which, with the exception
of a few languages, are not readily available.

In this paper we present a new linguistically uninformed model to
automatically label texts for animacy. We show that we can (mostly)
do away with features that require syntactic parsing or semantic
lexicons while still yielding competitive performance. We focus
on labeling animacy in stories because stories pose some interesting
problems to automatic systems of animacy recognition. As the example
of the fleeing pancake already made clear, quite some entities in
stories, though inanimate in the `real' world, exhibit all behavioral
capacities to consider them animate and rational agents. Such
examples make a clear case for developing dynamic, data driven systems
that do not rely too much on static and fixed world knowledge.

@@

\cite{ovrelid:04,ovrelid:05,ovrelid:06,ovrelid:08,ovrelid:09} present
a number of animacy classifiers that make use of syntactic and
morphological features. These features include the frequency of
analysis of the noun as `subject' or `object', the frequency of the
occurence of a noun in a passive $by$-phrase, and the frequency of the
noun as a subject followed by either animate personal pronouns or
inanimate personal pronouns. These features are then aggregated for
each lemma after which a machine learning system (decision tree or
$k$-nearest neighbor classifier) is trained. In \cite{bowman:12} a
similar approach is presented. In this study a Maximum Entropy
classifier is trained on the basis of three feature types: (1)
bag-of-words with and without their corresponding Part-of-Speech tags,
(2) internal syntactic features such as the syntactic head and (3)
external syntactic features that describe the dependency relation of a
noun to a verb (i.e.\ subject relation, object relation etc.)  This is
the only study that makes use of a fully labeled corpus for animacy.
Partially related to animacy detection, \cite{karsdorp:12}
attempt to extract the cast (i.e.\ all characters) from a
story. Similar to \cite{bowman:12} they rely on dependency tags to
extract the subjects of direct and indirect speech.

@@

% commentary on above approaches
The approaches mentioned above present us with a number of
problems. First, almost all of them heavily rely on costly,
linguistically informed features derived from lexical-semantic
databases or syntactic parsing. For most languages in the world,
however, we cannot rely on these resources, either because they do not
exist, or because their performance is insufficient.  Second, animacy
detection is often seen as a useful feature for a range of natural
language processing techniques, such as anaphora resolution and
syntactic parsing. The mutual dependence between these techniques and
animacy detection, however, leads to a chicken and egg
situation. Another problem with the approaches above is that they are
lemma-based, which means that the models are generally insensitive to
different usages of a particular word. A pancake is typically not an
animate entity, but, as we have seen, whether it is animate or not may
depend on the story in which it appears, and even within the story the
animacy may change. We therefore need to develop a dynamic, data
driven system that is token based and does not rely too much on static
and fixed world knowledge.

@@

\subsubsection{Semantic Features} The most innovative feature we have
included in our model is concerned with semantic similarity. In his
\textit{Philophische Untersuchungen} Wittgenstein already suggests
that ``Die Bedeutung eines Wortes ist sein Gebrauch in der Sprache''
(PI 43). The well-known insight in computational linguistics that the
meaning of words can be approximated by comparing the linguistic
contexts in which words appear, is an implementation of this very
idea. The idea is very simple, yet powerful: words that often
co-appear with the same set of words, will have a more similar
meaning. Recently, there has been a lot of interest in procedures that
can automatically induce so-called `word embeddings' from large,
unannotated collections of texts. These models typically attempt to
learn a lower dimensional vector for each word in the vocabulary which
captures the typical co-occurrence patterns of the word in the
corpus. The similarity between words can then be approximated by
applying arithmetics to these vectors of word embeddings.

@@

We add the word embedding vectors to our animacy model as follows. We
augment each word in the data with its corresponding 300 dimension
word embeddings vector. Out-of-vocabulary words were given a vector in
which all dimensions are set to zero.

@@

In Table \ref{tab:results-all} we present the results for all nine
models on the complete data set. For each model we report the
precision, recall, $F1$-score and Area Under the Precision-Recall
Curve (AUC) for the animate words and the
inanimate words.

\begin{table}
\centering
\begin{tabular}{llrrrr}
\toprule
features & class & precision &  recall & $F1$ &  AUC \\
\midrule
\multirow{2}{*}{embeddings}                            & inanimate     &       0.98 &    0.99 &    0.98 & \multirow{2}{*}{0.93} \\
                                                       & animate     &       0.92 &    0.89 &    0.90 &  \\
\multirow{2}{*}{word}                                  & inanimate     &       0.96 &    0.99 &    0.98 & \multirow{2}{*}{0.93} \\
                                                       & animate     &       0.94 &    0.78 &    0.85 &  \\
\multirow{2}{*}{word + embeddings}                     & inanimate     &       0.98 &    0.99 &    0.98 & \multirow{2}{*}{0.96} \\
                                                       & animate     &       0.93 &    0.90 &    0.91 &  \\
\multirow{2}{*}{word + PoS}                            & inanimate     &       0.97 &    0.99 &    0.98 & \multirow{2}{*}{0.95} \\
                                                       & animate     &       0.94 &    0.86 &    0.89 &  \\
\rowcolor{Gray}                                        & inanimate     &       0.98 &    0.99 &    0.99 &  \\
\rowcolor{Gray}\multirow{-2}{*}{word + PoS + embeddings}   & animate     &       0.94 &    0.91 &    0.93 & \multirow{-2}{*}{0.97} \\
\multirow{2}{*}{word + PoS + lemma}                    & inanimate    &       0.97 &    0.99 &    0.98 & \multirow{2}{*}{0.96} \\
                                                       & animate     &       0.94 &    0.86 &    0.90 &  \\
\multirow{2}{*}{word + PoS + lemma + embeddings}       & inanimate     &       0.98 &    0.99 &    0.99 & \multirow{2}{*}{0.97} \\
                                                       & animate     &       0.94 &    0.91 &    0.93 &  \\
\multirow{2}{*}{word + PoS + lemma + dep}              & inanimate     &       0.97 &    0.99 &    0.98 & \multirow{2}{*}{0.96} \\
                                                       & animate     &       0.94 &    0.86 &    0.90 &  \\
\multirow{2}{*}{word + PoS + lemma + dep + embeddings} & inanimate     &       0.98 &    0.99 &    0.99 & \multirow{2}{*}{0.97} \\
                                                       & animate     &       0.94 &    0.92 &    0.93 &  \\
\bottomrule
\end{tabular}
\caption{Precision, recall, $F1$ and Area Under the Curve (AUC) scores for animate and inanimate classes per feature setting for all words.}
\label{tab:results-all}
\end{table}

All models perform well on classifying inanimate words. However, since
this is the majority class, it is of course far more interesting to
compare the performance of the models on the animate instances. It is
interesting to observe that the `simple' $n$-gram word model already
performs rather well. Adding more features, such as Part-of-Speech,
lemmata, etcetera, only has a positive impact on the recall of the
model, while leaving the precision untouched. As can be observed from
the table, employing the rather expensive dependency features shows
barely any improvement.

@@

\begin{table}
\centering
\begin{tabular}{llrrrr}
\toprule
features & class &  precision &  recall & $F1$ &  AUC \\
\midrule
\multirow{2}{*}{embeddings}                            & inanimate     &       0.89 &    0.94 &    0.92 & \multirow{2}{*}{0.92} \\
                                                       & animate       &       0.92 &    0.84 &    0.88 &  \\
\multirow{2}{*}{word}                                  & inanimate     &       0.78 &    0.98 &    0.87 & \multirow{2}{*}{0.91} \\
                                                       & animate       &       0.96 &    0.60 &    0.74 &  \\
\multirow{2}{*}{word + embeddings}                     & inanimate     &       0.90 &    0.96 &    0.93 & \multirow{2}{*}{0.96} \\
                                                       & animate       &       0.94 &    0.86 &    0.89 &  \\
\multirow{2}{*}{word + pos}                            & inanimate     &       0.86 &    0.96 &    0.90 & \multirow{2}{*}{0.94} \\
                                                       & animate       &       0.93 &    0.78 &    0.84 &  \\
\rowcolor{Gray}                                        & inanimate     &       0.93 &    0.96 &    0.94 &  \\
\rowcolor{Gray}\multirow{-2}{*}{word + PoS + embeddings} & animate       &       0.94 &    0.89 &    0.92 &\multirow{-2}{*}{0.97}  \\
\multirow{2}{*}{word + PoS + lemma}                    & inanimate     &       0.87 &    0.96 &    0.91 & \multirow{2}{*}{0.95} \\
                                                       & animate       &       0.94 &    0.80 &    0.86 &  \\
\multirow{2}{*}{word + PoS + lemma + embeddings}       & inanimate     &       0.92 &    0.96 &    0.94 & \multirow{2}{*}{0.97} \\
                                                       & animate       &       0.94 &    0.89 &    0.92 &  \\
\multirow{2}{*}{word + PoS + lemma + dep}              & inanimate     &       0.87 &    0.96 &    0.91 & \multirow{2}{*}{0.95} \\
                                                       & animate       &       0.93 &    0.80 &    0.86 &  \\
\multirow{2}{*}{word + PoS + lemma + dep + embeddings} & inanimate     &       0.93 &    0.96 &    0.94 & \multirow{2}{*}{0.97} \\
                                                       & animate       &       0.94 &    0.89 &    0.92 &  \\
\bottomrule
\end{tabular}
\caption{Precision, recall, $F1$ and Area Under the Curve (AUC) scores for animate and inanimate classes
  per feature settings for all words tagged as noun.}
\label{tab:results-noun}
\end{table}
